\section{Discussion and Conclusion}
\label{sec:discussion}

\paragraph{Summary.}
This paper makes two contributions. First, we identify and empirically validate a fundamental failure mode of Lagrangian pruning in MoE architectures: per-neuron FLOPs penalty gradients are ${\sim}100{\times}$ weaker than task gradients, causing uniform mask collapse rather than selective pruning. The Lagrangian baseline confirms this prediction catastrophically on all three datasets ($10.4\%$ on CIFAR-100, $36.4\%$ on CIFAR-10, $4.7\%$ on Tiny-ImageNet) despite reaching the target FLOPs ratio---a result that should inform future work on MoE compression. Second, we show that a simple fix---replacing the learned penalty with a deterministic cubic sparsity schedule, gradient masking, and dual-rate optimization (KaleidoNet)---restores effective pruning: KaleidoNet retains 98.3--99.2\% of dense-model accuracy with $1.8{\times}$ FLOPs reduction across all three datasets, consistently outperforming all pruning baselines.

\paragraph{What is novel.}
The novelty of this work is the \emph{diagnosis}, not the architecture. The gradient-scale mismatch we identify (Section~\ref{sec:method}) is, to our knowledge, the first formal explanation of why per-neuron Lagrangian FLOPs pruning collapses inside MoE sub-networks, and the empirical sweep across four pruning baselines isolates that collapse to the gradient signal itself: every method that supplies the budget signal externally (Linear, Random, Magnitude, and our cubic variant) bypasses the failure mode, while the only method that supplies it as a Lagrangian gradient (the Lagrangian baseline) collapses on every dataset. KaleidoNet is best understood as a stable, well-tuned instantiation of the schedule-based remedy---the cubic schedule, gradient masking, and dual-rate optimizer are useful engineering choices for practitioners, but they are secondary to the underlying insight. Readers seeking to apply this work should treat the diagnosis as the contribution that transfers across MoE compression settings, with KaleidoNet as a starting recipe rather than the central artifact.

\paragraph{Scope: FLOPs vs.\ wall-clock.}
The compression claims of this work are scoped to FLOPs and parameter granularity, not wall-clock latency. We report a $1.8{\times}$ FLOPs reduction and a $1.63{\times}$ parameter compression after model surgery, both of which are properties of the trained network and are reproducible from the released weights. Wall-clock latency is a separate question that depends on the inference stack: on the Apple MPS platform we benchmarked, MoE routing overhead dominates at this 5.49M-parameter scale and KaleidoNet runs ${\sim}18{\times}$ slower than the dense ViT (Table~\ref{tab:latency}), which we report as a known platform-specific limitation rather than a property of the schedule-based approach. Closing the wall-clock gap would require larger expert widths and CUDA-fused MoE kernels (e.g.\ MegaBlocks); we have not benchmarked these in this paper, and we explicitly do not claim a deployment speedup. The compression result is what generalizes; the latency result is platform-bound.

\paragraph{Limitations.}
Several limitations should be noted:
\begin{itemize}
    \item \textbf{Dense model still wins on accuracy alone.} The unpruned dense model outperforms KaleidoNet on all three datasets (by 0.5--1.1 pp). KaleidoNet's advantage is in the accuracy--efficiency tradeoff (98.3--99.2\% of dense accuracy at $1.8{\times}$ fewer FLOPs), not in absolute accuracy. When compute is unconstrained, the dense model remains the better choice.
    \item \textbf{Scale.} Our converged experiments span three datasets (CIFAR-10, CIFAR-100, Tiny-ImageNet) at 32$\times$32 resolution with up to 200 classes. Validation on ImageNet-1K at full resolution would further strengthen generality claims but requires GPU cluster resources beyond our current setup.
    \item \textbf{Attention head pruning.} The cubic schedule currently targets only ElasticLinear neuron widths. All 6 attention heads survive pruning, leaving potential compression on the table.
\end{itemize}

\paragraph{Future work.}
Extending the cubic schedule to jointly prune attention heads and expert widths, scaling to ImageNet-1K at full resolution, and investigating the interplay between expert count and pruning depth are promising research directions. The framework is also applicable to language models, where MoE layers are widely used.

\paragraph{Broader impact.}
Structured pruning methods such as KaleidoNet contribute to reducing the computational and energy footprint of neural network deployment. As vision transformers are increasingly adopted in real-world applications---from medical imaging to autonomous driving---the ability to compress models without specialized sparse hardware has direct implications for deployment on resource-constrained edge devices (mobile phones, embedded systems, IoT sensors). Smaller models require less memory, fewer FLOPs, and consequently less energy per inference, contributing to reduced carbon emissions at scale. Democratizing access to efficient models can also lower the barrier for researchers and practitioners in compute-limited settings. On the other hand, more efficient models may accelerate deployment in surveillance or other sensitive applications. We do not foresee negative societal impacts specific to our pruning methodology, but we encourage responsible deployment practices for any compressed model.
