% Why Lagrangian Pruning Fails in Mixture-of-Experts:
% Gradient-Scale Mismatch and a Schedule-Based Fix
% Target venue: TMLR or NeurIPS 2026
%
% Options: [final] for camera-ready, [preprint] for arXiv, default = submission
\documentclass[10pt]{article}
\usepackage{neurips_2026}

% ---- Extra packages ----
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{nicefrac}

% ---- Math shortcuts ----
\newcommand{\R}{\mathbb{R}}
\newcommand{\bm}{\mathbf{m}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\btheta}{\boldsymbol{\theta}}
\DeclareMathOperator{\softmax}{softmax}
\DeclareMathOperator{\sigmoid}{sigmoid}

% ---- Theorem environments ----
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

% ====================================================================
\title{Why Lagrangian Pruning Fails in Mixture-of-Experts:\\Gradient-Scale Mismatch and a Schedule-Based Fix}

\author{
  Karim Magdy$^{1(\boxtimes)}$\thanks{Corresponding author.} \\
  \And
  Ghada Khoriba$^{1,2}$ \\
  \And
  Mohamed N. Swailam$^{4}$ \\
  \And
  Hala Abbas$^{1,3}$ \\
  \\
  $^1$Computer Science Department, Faculty of Computers and Artificial Intelligence, \\
  Helwan University, Helwan, Egypt \\
  \texttt{\{karimmagdy\_csp,ghada\_khoriba\}@fci.helwan.edu.eg} \\
  $^2$School of Information Technology and Computer Science (ITCS), \\
  Nile University, Giza, Egypt \\
  \texttt{ghadaKhoriba@nu.edu.eg} \\
  $^3$Faculty of Computer Studies, Arab Open University, Cairo, Egypt \\
  \texttt{hala.abbas@aou.edu.eg} \\
  $^4$Department of Informatics, Systems and Communication, \\
  Universit\`a degli Studi di Milano-Bicocca, Milano, Italy \\
  \texttt{m.sweilam@campus.unimib.it}
}

\begin{document}
\maketitle

% ====================================================================
\input{sections/abstract}

\input{sections/introduction}

\input{sections/related_work}

\input{sections/method}

\input{sections/experiments}

\input{sections/discussion}

% ====================================================================
\bibliographystyle{plainnat}
\bibliography{references}

% ====================================================================
% Appendix (optional, unlimited pages for NeurIPS)
\newpage
\appendix
\input{sections/appendix}

% ====================================================================
% NeurIPS Paper Checklist
\newcommand{\answerYes}[1][]{\textcolor{blue}{[Yes]}~#1}
\newcommand{\answerNo}[1][]{\textcolor{orange}{[No]}~#1}
\newcommand{\answerNA}[1][]{\textcolor{gray}{[N/A]}~#1}

\section*{NeurIPS Paper Checklist}
\begin{enumerate}
\item {\bf Claims}
\item[] Question: Do the main claims accurately reflect the paper's contributions?
\item[] Answer: \answerYes{}
\item[] Justification: All claims are supported by converged experiments across 3 datasets (CIFAR-10, CIFAR-100, Tiny-ImageNet) with 3 seeds each and 6 methods including the Lagrangian baseline.

\item {\bf Limitations}
\item[] Question: Does the paper discuss limitations?
\item[] Answer: \answerYes{}
\item[] Justification: Section~6 discusses dense model superiority on accuracy alone, scale limitations (no ImageNet-1K), attention head pruning gap, and 18$\times$ inference overhead.

\item {\bf Theory Assumptions and Proofs}
\item[] Question: For each theoretical result, does the paper provide full assumptions and proof?
\item[] Answer: \answerYes{}
\item[] Justification: The gradient-scale mismatch analysis (Section~3) provides complete derivation with stated assumptions.

\item {\bf Experimental Result Reproducibility}
\item[] Question: Does the paper disclose all information needed to reproduce results?
\item[] Answer: \answerYes{}
\item[] Justification: All hyperparameters in Section~5.1 and Appendix. Algorithm~1 provides pseudocode. Code available at \url{https://github.com/karimmagdy/KaleidoNet}.

\item {\bf Open access to data and code}
\item[] Question: Does the paper provide open access to data and code?
\item[] Answer: \answerYes{}
\item[] Justification: All datasets are public benchmarks. Code is open-source at \url{https://github.com/karimmagdy/KaleidoNet}.

\item {\bf Experimental Setting/Details}
\item[] Question: Does the paper specify all training and test details?
\item[] Answer: \answerYes{}
\item[] Justification: Section~5.1 specifies architecture, optimizer, learning rates, schedules, and evaluation protocol.

\item {\bf Experiment Statistical Significance}
\item[] Question: Does the paper report error bars or statistical significance?
\item[] Answer: \answerYes{}
\item[] Justification: All results report mean $\pm$ std over 3 seeds. Per-seed results in Appendix Tables A1--A3.

\item {\bf Experiments Compute Resources}
\item[] Question: Does the paper provide compute resource information?
\item[] Answer: \answerYes{}
\item[] Justification: Section~5.6 reports inference latency and training time. Appendix provides per-configuration compute.

\item {\bf Code Of Ethics}
\item[] Answer: \answerYes{}

\item {\bf Broader Impacts}
\item[] Answer: \answerYes{}
\item[] Justification: Section~6 discusses computational efficiency benefits, edge deployment implications, and potential misuse in surveillance.

\item {\bf Safeguards}
\item[] Answer: \answerNA{}
\item[] Justification: The method is a general pruning algorithm with no specific misuse risks.

\item {\bf Licenses}
\item[] Answer: \answerYes{}
\item[] Justification: All datasets and baseline methods are properly cited.

\item {\bf New Assets}
\item[] Answer: \answerYes{}
\item[] Justification: KaleidoNet is fully described in Algorithm~1 and the method section. Code is open-source.

\item {\bf Human Subjects}
\item[] Answer: \answerNo{}

\item {\bf IRB Approvals}
\item[] Answer: \answerNA{}
\end{enumerate}

\end{document}
